Approximation as Differentiation

### 2025-06-07 main/first-order.tex
Pare back LatGal a bit.

@@

\todo{Rework to account for new ordering.} Galois connections are pairs of monotone functions $f: Y \to X$ and
$g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse of $g$,
and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the bidirectional
nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the least input
slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the greatest
output slice for a given slice of the input).

@@

\noindent Galois connections thus generalise order isomorphisms. The $\adj$ notation is justified because a
Galois connection $f \adj g: X \to Y$ can also be seen an adjunction between poset categories, with monotone
$f$ and $g$ regarded as functors; $f$ is usually referred to as the \emph{upper} (right) adjoint and $g$ as
the \emph{lower} (left) adjoint. Galois connections compose component-wise, with $\id_X \adj \id_X: X \to X$
as the unit for composition, and thus form a category $\PosGal$ with all posets as objects and all Galois
connections between them as morphisms.

As sketched in \secref{introduction:galois-slicing}, for \GPS we would like a setting where the approximants
of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing
the approximation that discards all information about $x$, greatest element $\top$ representing the
approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to
combine approximations. Thus rather than working directly in $\PosGal$, we will use the following subcategory
instead.

@@

$\LatGal$ is not Cartesian closed, and so not directly suitable for interpreting higher-order programs. In
fact Cartesian closure is incompatible with another important property of $\LatGal$. The projections $\pi_1: X
\times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the product of lattices, have both
upper and lower adjoints; this means that $X \times Y$, which we shall hereafter write as $X \biprod Y$, is
both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$ and injections $\biinj_X$ and
$\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and so $\cat{C}$ has only a single object, up to isomorphism.

>>>>

A biproduct category $\cat{C}$ cannot be Cartesian closed without being trivial. If $0 \times X \iso X$ then
$\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X \Rightarrow Y} \iso 1$, and so $\cat{C}$ is terminal, with only a
single object up to isomorphism.

@@

\todo{Rework to account for new ordering.} Galois connections are pairs of monotone functions $f: Y \to X$ and
$g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse of $g$,
and $g$ the best approximation from above to an inverse of $f$. The category $\LatGal$ of bounded lattices and
Galois connections is semi-additive.

@@

\noindent The $\adj$ notation is justified because a Galois connection $f \adj g: X \to Y$ can also be seen an
adjunction between poset categories, with monotone $f$ and $g$ regarded as functors; $f$ is usually referred
to as the \emph{upper} (right) adjoint and $g$ as the \emph{lower} (left) adjoint. Galois connections compose
component-wise, with $\id_X \adj \id_X: X \to X$ as the unit for composition.

@@

The projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the
product of lattices, have both upper and lower adjoints; this means that $X \times Y$, which we shall
hereafter write as $X \biprod Y$, is both a product and a coproduct, with projections $\biproj_X$ and
$\biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$.
